%!XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
%!XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
%!X
%!X     libAtoms: atomistic simulation library
%!X     
%!X     Copyright 2006-2007.
%!X
%!X     Authors: Gabor Csanyi, Steven Winfield, James Kermode
%!X     Contributors: Noam Bernstein, Alessio Comisso
%!X
%!X     The source code is released under the GNU General Public License,
%!X     version 2, http://www.gnu.org/copyleft/gpl.html
%!X
%!X     If you would like to license the source code under different terms,
%!X     please contact Gabor Csanyi, gabor@csanyi.net
%!X
%!X     When using this software, please cite the following reference:
%!X
%!X     http://www.libatoms.org
%!X
%!XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
%!XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX

\section{Copyright}

\textsc{libAtoms} is copyright 2006-2008.

\begin{description}
\item[Authors:] Gabor Csanyi, Steven Winfield, James Kermode
\item[Contributors:] Noam Bernstein, Alessio Comisso, Lars Pastewka
\end{description}

The source code is released under the GNU General Public License,
version 2, \\ \texttt{http://www.gnu.org/copyleft/gpl.html}.

If you would like to license the source code under different terms,
please contact Gabor Csanyi, \texttt{gabor@csanyi.net}.

When using this software, please cite the following reference:
\texttt{http://www.libatoms.org}

\section{Introduction}

\subsection*{Purpose and Philosophy}

\textsc{libAtoms}\footnote{http://www.libatoms.org} is a Fortran 95
library for developing molecular dynamics codes. It has been designed
to be extensible and efficient, and can be used both serially and on
large parallel computers that offer an implementation of the MPI
library. At present the parallisation is at a very early stage, and
there is no parallelisation of data. The fundamental design principal
behind the developement of \textsc{libAtoms} is the idea of
\emph{expressive programming} --- this involves making sensible use of
high level language features to produce clear and easily understood
code that is as close as possible to the abstract algorithm that is
being implemented.

\textsc{libAtoms} is structured as a series of Fortran modules, each of which
can be thought as self contained entities rather like C++ classes. Most
modules define a type that takes its name from that of the module
(for example the `Atoms' type in the module \texttt{Atoms\_module}),
and a series of subroutines and functions that operate on objects
of that type, in analogue with the methods of a C++ class.

There are modules within \textsc{libAtoms} to represent a set of
atoms, to apply constraints to those atoms and to integrate their
equations of motion. There are also utility modules to perform task
such as reading parameters from files, to solve multi-dimensional
optimisation problems or to calculate the elastic properties of a
given force model. It is intended that users write their own top-level
programs which define instances of the various \textsc{libAtoms} types
and call \textsc{libAtoms} methods.

\subsection*{Requirements}

In order to facilitate the expressive programming ideas outlined
above, \textsc{libAtoms} requires some extensions to the core Fortran
95 standard. These extensions are all part of Fortran 2003, although
there are currenly no complete Fortran 2003 compilers available. They
are also all part of the ISO ``Allocatable Technical Report''
standard.\footnote{ISO TR-15581: Enhanced Data Type Facilities}, which
has been implemented by a number of modern compiler vendors. The
extensions required are:
\begin{itemize}
\item Allocatable components within derived types
\item Allocatable dummy arguments
\end{itemize}

At the time of writing, these features are known to be
supported by the following compilers:
\begin{itemize}
\item Intel \texttt{ifort} version 8.1 or later
\item \texttt{g95} version 0.91 or later
\item Pathscale \texttt{pathf90} version 2.5 or later
\item Portland group \texttt{pgf90} version ?? or later
\item Sun \texttt{f95} version 8.3 or later
\item HP Fortran V5.6 or later
\end{itemize}

\textsc{libAtoms} requires \textsc{lapack} and \textsc{blas}
libraries. It has been tested using the Intel \textsc{mkl} library,
but should work with any vendor supplied library.

If you want to use \textsc{libAtoms} on a parallel machine, you need
an MPI library for that architecture with Fortran bindings (i.e. with
a mpif.h header file). libAtoms has been tested with \textsc{mpich},
\textsc{lam} and Pathscale MPI, but should work with any standards complying
MPI implementation.

To generate the \textsc{libAtoms} documentation, Python 2.4 or later
 and a working \LaTeX{} installation are required. If you want the HTML
documentation then you will also need \LaTeX 2\texttt{HTML}.

\newpage

\subsection*{Example Program --- Classical Molecular Dynamics}

The best way to get a feel for how \textsc{libAtoms} is used
is with a simple example. The code in \texttt{libAtoms/md.f95} sets up
the \textsc{libAtoms} environment, loads a cell from
an XYZ file, and performs 1,000 steps of classical molecular
dynamics. For this example random forces are used, but in general
they would be calculated using a classical potential.

\subsection*{Controlling Output Verbosity}

Often it is useful to be able to vary the amount and detail of text
produced by a program, for example to get more detailed information
when debugging. \textsc{libAtoms} makes use of a global verbosity
stack to control the amount of output produced by its subroutines and
functions. The possible values, in increasing order, of verbosity are:
\texttt{ERROR=-10}, \texttt{SILENT=-1}, \texttt{NORMAL=0},
\texttt{VERBOSE=1}, \texttt{NERD=2} and \texttt{ANAL=3}.  Printing
interfaces for objects normally print the their own structure at some 
level of detail, and contained substructures at a lower level of detail.  
Other subroutines normally print their output at some detail level, and
request lower level subroutines to print at a lower detail level.
You can set a verbosity minimum level that does not allow subroutines
to lower the verbosity level, so that lower level routines and structures
always report at the requested detail level or higher.

The initial value at the top of the verbosity stack is set by
\texttt{system\_initialise()} (the default is \texttt{NORMAL} but this
can be overidden with an optional argument). The stack can be modified
by calling \texttt{push\_verbosity(val)}, \texttt{push\_verbosity\_increment(n)},
or \texttt{push\_verbosity\_decrement(n)} to place a new value on the top of the
stack, and \texttt{pop\_verbosity()} to revert to the previous value. 
The current value of the verbosity stack can be queried with 
\texttt{current\_verbosity()}.
The initial verbosity cascade status is set to false, and can be changed
with \texttt{verbosity\_set\_minimum()} and revert to the previous value
with \texttt{verbosity\_unset\_minimum()}.

Here are some examples:

\begin{boxedminipage}{\textwidth}
\begin{verbatim}
   type(Atoms) :: myatoms

   call system_initialise(NORMAL)
   ...
   call print(myatoms)                 ! Prints Atoms object
   call push(verbosity_stack,VERBOSE)  ! Temporarily switch to VERBOSE output
   call print(myatoms)                 ! Now this will be printed

   call pop(verbosity_stack)           ! Revert to NORMAL output level
   call verbosity_set_minimum(VERBOSE) ! force lower level things to output at 
                                       !  least at VERBOSE detail
   call print(myatoms)                 ! Now this will be printed at VERBOSE, 
                                       !  and subobjects too
   call verbosity_unset_minimum()      ! revert to previous verbosity cascade

   call system_finalise
\end{verbatim}
\end{boxedminipage}

This behaviour is implemented for printing interfaces as in this example

\begin{boxedminipage}{\textwidth}
\begin{verbatim}
  subroutine structure_print(this, out)
  type(Structure), intent(in) :: this
  type(inoutput), intent(inout), optional :: out

  call print(this%some_int, out)
  call verbosity_decrement() ! print following structure in a bit less detail
  call print(this%some_substructure, out = out)
  call verbosity_pop()

  call verbosity_decrement(2) ! print following structure in a lot less detail
  call print(this%other_substructure, out = out)
  call verbosity_pop()
\end{verbatim}
\end{boxedminipage}

Printing of low level entities (strings, numbers, arrays) take an optional
verbosity parameter that sets the minimum verbosity level at which that
print will actually occur.  Functions that want to produce output at
different levels of detail (e.g. the \texttt{minim()} function) can use
this verbosity parameter to output more information when the verbosity
level is high.  To modify output of a structure, the verbosity stack
must be used.  For example:

\begin{boxedminipage}{\textwidth}
\begin{verbatim}
  subroutine sub()
    ! do some stuff

    ! usually print this info
    call print("got value " // n, NORMAL) 

    ! only print when current level is set to verbose
    call print("extra internal value " // internal, VERBOSE) 

    ! print this structure at normal level of detail
    call print(struct)

    ! print this structure at less than usual detail, i.e.
    ! SILENT if current level is NORMAL
    call verbosity_push_decrement()
    call print(other_struct)
    call verbosity_pop()

    ! print this structure at a lot more than usual detail, i.e.
    ! NERD if current level is NORMAL
    call verbosity_push_increment(2)
    call print(other_struct)
    call verbosity_pop()

\end{verbatim}
\end{boxedminipage}

These controls can be used from upper level programs as follows:

\begin{boxedminipage}{\textwidth}
\begin{verbatim}
   call system_initialise(NORMAL)

   ...
   ! (1) Produce all the 'NORMAL' level output
   steps = minim(x,func,dfunc,'cg','LINMIN_FAST',1e-7_dp,500)

   ! (2) Print more detail (VERBOSE = 1)
   call verbosity_push(VERBOSE)
   steps = minim(x,func,dfunc,'cg','LINMIN_FAST',1e-7_dp,500)
   call verbosity_pop()

   ! (3) Print less detail (SILENT = -1)
   call verbosity_push(SILENT)
   steps = minim(x,func,dfunc,'cg','LINMIN_FAST',1e-7_dp,500)
   call verbosity_pop()

   ! (4) Same as 2, but lower level subroutines such as linmin will 
   !     also produce verbose output
   call verbosity_set_minimum(VERBOSE)
   steps = minim(x,func,dfunc,'cg','LINMIN_FAST',1e-7_dp,500)
   call verbosity_unset_minimum()

   call system_finalise
\end{verbatim}
\end{boxedminipage}
